---
title: "Homework 1"
author: "3230104912 Zhai Guoqing"
date: \today
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
    
---

```{r setup, message = F, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(DAAG)
```

# 1 
## a.Obtain the data frame iowa.df by running the following code:

```{r}
iowa.df<-read.csv("data/Iowa.csv", sep = ';', header=T)
```

## b.How many rows and columns does iowa.df have?

```{r}
dim(iowa.df)
```

## c. What are the names of the columns of iowa.df?

```{r}
colnames(iowa.df)
```

## d. What is the value of row 5, column 7 of iowa.df?

```{r}
iowa.df[5, 7]
```

## e. Display the second row of `iowa.df` in its entirety.

```{r}
iowa.df[2, ]  
```

# 2 
## a 
In R, anything inside quotes is treated as text (strings).

Since vector1 is of character data type, the max() function will compare values based on Unicode encoding and return the result. For "7" and "32", it first compares "7" and "3", therefore it should output "7".

The sorting rule for the sort() function is similar - it should output "12" "32" "5" "7".

The sum() function will generate an error because it expects numeric input rather than a character vector, but this can be resolved by converting to numeric using as.numeric().

## b 
```vector2 <- c("5",7,12)```

**Result**: Creates a character vector c("5","7","12").

```vector2[2] + vector2[3]```

**Result**: Error - "non-numeric argument to binary operator".

**Reason**: Both elements are character strings, so we cannot perform arithmetic operations on characters without conversion.

```dataframe3 <- data.frame(z1="5",z2=7,z3=12)```

**Result**: Creates a data frame with character "5", numeric 7 and numeric 12.

```dataframe3[1,2] + dataframe3[1,3]```

**Result**: 19.

```list4 <- list(z1="6", z2=42, z3="49", z4=126)```

**Result**: Creates a list with mixed types character "6", numeric 42, character "49" and numeric 126.

```list4[[2]]+list4[[4]]```

**Result**: 168.

```list4[2]+list4[4]``` 

**Result**: Error- "non-numeric argument to binary operator".

**Reason**: [ ] returns sublists but not the actual values.

# 3  
## a  
```{r}
seq(from = 1, to = 10000, by = 372)
```

```{r}
seq(from = 1, to = 10000, length.out = 50)
```

## b  
"rep(1:3, times=3)" repeats the entire vector 1:3 three times (1,2,3,1,2,3,1,2,3), while "rep(1:3, each=3)" repeats each element three times consecutively (1,1,1,2,2,2,3,3,3).

The key difference is repetition granularity: "times" repeats the whole sequence, "each" repeats individual elements.

# MB.Ch1.2
```{r}
library(DAAG)
data(orings)

selected_rows <- c(1, 2, 4, 11, 13, 18)
new_orings <- orings[selected_rows, ] 

par(mfrow = c(1, 2), mar = c(5, 4, 4, 2) + 0.1)

plot(new_orings$Temperature, new_orings$Total,
     xlim = c(50, 85), ylim = c(0, 6),
     xlab = "Temperature (F)", ylab = "Total Incidents",
     main = "Pre-Challenger Data (Subset)",
     pch = 19, col = "red", cex = 1.5)
grid()

plot(orings$Temperature, orings$Total,
     xlim = c(50, 85), ylim = c(0, 6),
     xlab = "Temperature (F)", ylab = "Total Incidents",
     main = "Complete O-Ring Data",
     pch = 21, bg = "blue", cex = 1.2)
points(new_orings$Temperature, new_orings$Total, 
       pch = 19, col = "red", cex = 1.5)
grid()
legend("topright", legend = c("Subset", "Full Data"),
       pch = c(19, 21), col = c("red", "black"),
       pt.bg = c(NA, "blue"))
```

# MB.Ch1.4
## a
```{r}
library(DAAG)
data(ais)

str(ais)
missing_values <- colSums(is.na(ais))
print(missing_values)
```

There are no missing values.

## b
```{r}
gender_sport <- table(ais$sex, ais$sport)
colnames(gender_sport) <- c("B_ball", "Field", "Gym", "Netball", "Row", "Swim", "T_400m", "T_sprnt", "Tennis", "W_Polo")

print(gender_sport)
imbalance_threshold <- 2
gender_ratio <- gender_sport["f", ] / gender_sport["m", ]
imbalanced_sports <- names(which(gender_ratio > imbalance_threshold|
                              gender_ratio < 1/imbalance_threshold))
print(imbalanced_sports)
```

# MB.Ch1.6
## a
```{r}
attach(Manitoba.lakes)
plot(log2(area) ~ elevation, pch=16, xlim=c(170,280))
# NB: Doubling the area increases log2(area) by 1.0
text(log2(area) ~ elevation, labels=row.names(Manitoba.lakes), pos=4)
text(log2(area) ~ elevation, labels=area, pos=2)
title("Manitoba's Largest Lakes")
```

Caption:
Y-axis (log₂(area))：Uses a base-2 logarithmic scale to represent lake area (in km²). A 1-unit increase means the area doubles.

Point: Right-side text represents lake name. Left-side number represents actual area.

A 1-unit gap = 2× area difference.

## b
```{r}
plot(area ~ elevation, pch=16, xlim=c(170,280), ylog=T)
text(area ~ elevation, labels=row.names(Manitoba.lakes), pos=4, ylog=T)
text(area ~ elevation, labels=area, pos=2, ylog=T)
title("Manitoba's Largest Lakes")
```

# MB.Ch1.7
```{r}
dotchart(Manitoba.lakes$area,
         labels = row.names(Manitoba.lakes),
         xlab = "Area (km²)",
         main = "Manitoba Lakes (Linear Scale)",
         pch = 16)
dotchart(log2(Manitoba.lakes$area),
         labels = row.names(Manitoba.lakes),
         xlab = "log2(Area)",
         main = "Manitoba Lakes (Logarithmic Scale)",
         pch = 16)
abline(v = seq(10, 15, by=1), col="gray", lty=2)
# Optional: Add actual area values on the right side
text(x = log2(Manitoba.lakes$area), 
     y = 1:nrow(Manitoba.lakes),
     labels = Manitoba.lakes$area,
     pos = 1, cex = 0.8)
```

# MB.Ch1.8
```{r}
total_water_area <- sum(Manitoba.lakes$area)
print(total_water_area)